\documentclass{article}
\usepackage[final]{neurips}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{longtable}

% Adjust list indentation
\setlist{leftmargin=*}
\setlist[enumerate]{label=\arabic*., topsep=0pt, parsep=0pt}
\setlist[itemize]{topsep=0pt, parsep=0pt}

% Table settings
\setlength{\LTcapwidth}{\textwidth}
\setlength{\tabcolsep}{4pt}

% NeurIPS settings
% Camera-ready version is set in document class options

% Title and author information
\title{Co-aligning AI and Institutions: A Framework for Full-Stack Alignment}

\author{
  Joe Edelman\\
  Meaning Alignment\\
  % Add other authors similarly
}

\begin{document}

\maketitle

\begin{abstract}
AI alignment cannot be solved by focusing on a single system in isolation. Even perfectly intent-aligned AI embedded in misaligned institutions will lead to dangerous outcomes. The emerging field of ``sociotechnical alignment'' is currently undermined by reliance on two inadequate paradigms: the ``Standard Institution Design Toolkit'' (SIDT)---preference/utility frameworks from economics and game theory---and naive value-representations without structural guarantees. Both fail to account for moral reasoning, value evolution, and social context. Through five application areas---AI value stewardship, normatively competent agents, win-win negotiation systems, meaning-preserving economic mechanisms, and democratic regulatory institutions---we demonstrate why these thin models of human rationality cannot address our sociotechnical alignment challenges. We propose a new toolkit of thick models of choice that makes previously intractable problems tractable and enables co-alignment of AI and institutions toward global human flourishing. We call this ``Full-Stack Alignment'' (FSA), referring to the co-alignment of AI systems and institutions across the entire societal stack using thick models of choice that explicitly represent values and norms. Finally, we present a Full Stack Alignment Research Proposal for moving from theoretical foundations to practical adoption.
\end{abstract}

% Include all sections in the right order
\input{introduction}
\input{existing_toolkits}
\input{new_toolkit}
\input{case_studies}
\input{roadmap}
\input{conclusion}
\input{footnotes}
\input{related_work}

\renewcommand{\bibsection}{\section*{References}}
\setlength{\bibsep}{0.0pt}
\bibliographystyle{plainnat}
\bibliography{references}

\end{document}